---
title: "Practical: Introduction to Spatial Data"
author: "Kamarul Imran Musa"
date: "8/6/2019" 
output: 
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Workflow

- Prepare environment : 
  - directory
  - load packages
  - spatial data (Malaysia and Dengue)
- Read data 
  - polygons
  - points
- Analyze 

# Preparing environment for analysis

The packages for 

- running spatial analysis
- data wrangling
- import data
- deals with regular expressions

```{r}
library(sf)
library(tidyverse)
library(here)
library(janitor)
library(DT)
library(stringr)
library(readxl)
library(broom)
library(tmap)
library(mapview)
library(lubridate)
```

# Download data

1. Create a folder in `My Documents`
2. Name folder is `gis`
3. Download data (folder `data`)
4. Paste folder `data` inside folder `gis`  

# Create new RStudio project

1. File
2. New Project
3. Existing Directory
4. Browse for folder `gis`
 

# Read polygons data

```{r}
kel <- st_read(here("data", 
                    "Map_Kelantan",
                    "kelantan.shp")) 
kel %>% datatable()
```

The geometry is *Projected CRS: Kertau (RSO) / RSO Malaya (m)*

```{r}
st_geometry(kel)
```
Make plot

```{r}
plot(kel[,2]) #daerah
plot(kel[,3]) #mukim
plot(kel[,6]) #population
```

Or using geom_sf from **ggplot** package


```{r}
kel %>% 
  ggplot() +
  geom_sf(aes(fill = JUM_JANTIN)) +
  ggtitle('Population by subdistrict') +
  theme_bw()
```

Another way of plotting (by changing the color palette)

```{r}
ggplot() + 
  geom_sf(data = kel, aes(fill = JUM_JANTIN)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('Population by subdistrict') +
  theme_bw()
```

or using the **tmap** package

In a non-interactive mode

```{r}
tm_shape(kel) +
  tm_polygons("JUM_JANTIN")
```

Adding facets (by splitting the map based on gender)

```{r}
tm_shape(kel) +
  tm_polygons(c("LELAKI", "PEREMPUAN")) +
  tm_facets(ncol = 2)
```

Adding a variable for are using `sr_area()`

```{r}
kel <- kel %>% mutate(area_muk = st_area(kel),
                      area_muk_km2 = st_area(kel)/1000000)
kel %>% datatable()
```


# Dengue data

- This is vector data in points format.
- The dengue file comes in `xlsx` format. 
- So we will read using **read_xl** package 

```{r}
den <- read_xlsx(here('data',
                      'dengue_2019.xlsx')) %>%
  clean_names()
glimpse(den)
```

# Data cleaning

- select the first 8 letters of variable name

```{r}
den2 <- den %>% select_all(~str_sub(., end = 8L))
glimpse(den2)
```

Need to remove missing data (NA) in coordinate columns 

```{r}
den2 <- den2 %>% 
  filter(!is.na(latitud_),
         !is.na(lngitud_))
```

# Convert the dengue data to spatial data

- The excel is in WGS84 format so `CRS = 4326`. 
- Then let's describe `den` (now is spatial format)


```{r}
loc_den <- st_as_sf(den2, 
                    coords = c("lngitud_", "latitud_"), 
                    crs = 4326)
loc_den %>% datatable()
```

Let's confirm the `CRS` in WGS84 (EPSG:4326)

```{r}
st_crs(loc_den)
```

Make plots to see if map is alright (no outliers)

```{r}
ggplot() +
  geom_sf(data = loc_den) +
  ggtitle("Map of Dengue 2019") +
  theme_bw()
```


# Convert shapefile to RSO (https://epsg.io/3168)

Reasons

- the kel polygons is in RSO
- so the dengue points should also be in RSO

```{r}
loc_den2 <- st_transform(loc_den, 3168)
loc_den2 %>% datatable()
```

Now select points that fall only in Kelantan. Experince tells us there are always points outside the required boundary

```{r}
den_kel <- loc_den2 %>% 
  mutate(within_kel = lengths(st_within(loc_den2, kel)))
den_kel %>% datatable()
```

```{r}
den_kel <- den_kel %>% 
  filter(within_kel == 1)
```


# Dealing with dates

- variable `tarikh_o` and `tkh_noti` are in `chr` format
- need to change chr to the correct `dates` format
- https://lubridate.tidyverse.org/
- http://daniellequinn.github.io/RLessons/FormattingDates/FormattingDates.html 

```{r}
den_kel <- den_kel %>% 
  mutate_at(vars(tarikh_o, tkh_noti), ~dmy(.))
```

We can split dates into

- month
- day
- year


```{r}
den_kel <- den_kel %>% 
  mutate(year_on =  year(tarikh_o),
         month_on = month(tarikh_o),
         day_on = day(tarikh_o))
den_kel %>% datatable()
```


Generate epiweek

```{r}
den_kel <- den_kel %>% 
  mutate(epiweek_on = epiweek(tarikh_o))
```

Now we have

- polygon data
- point data

```{r}
glimpse(den_kel)
glimpse(loc_den2)
```


# Spatial plots

The overall plot for the year 2019

```{r}
den_kel <- den_kel %>% 
  filter(year_on == 2019) 
```

The plots

```{r}
den_plot <- ggplot() +
  geom_sf(data = kel) +
  geom_sf(data = den_kel) +
  ggtitle("Map of Dengue Cases for 2019") +
  theme_bw()
den_plot
```

Now let's make plots based on 

- jantina
- year of onset
- month of onset

```{r }
den_plot + 
  facet_wrap(~jantina)
```

```{r fig.width=6, fig.height=12}
den_plot + 
  facet_wrap(~ year_on + month_on, ncol = 3)
```

```{r fig.width=6, fig.height=12}
den_plot + 
  facet_wrap(~ DAERAH, ncol = 3)
```

Further analysis for year 2019

- number of dengue cases (column n) based on year, month and epid week
- bar chart for the total number per epi week
- line chart for the trend of dengue for each epid week


```{r}
# count den per epiweek
den_kel %>% 
  filter(year_on == 2019) %>% 
  group_by(year_on, month_on, epiweek_on) %>% 
  count() %>% 
  print(n = 52) %>%
  ggplot(aes(x = epiweek_on, y = n)) + 
  geom_bar(stat = 'identity', fill = 'white') +
  geom_line(colour = "red") +
  ggtitle('Number of dengue cases based on epid week') +
  ylab('Number') +
  xlab('Epid week based on date of onset') +
  theme_bw()
```

Using the **tmaps** package

All data for all years

```{r}
tm_shape(kel) +
  tm_polygons("DAERAH") + 
  tm_shape(den_kel) +
  tm_dots()
```

Split plots

```{r}
tm_shape(kel) +
  tm_polygons("NEGERI") +
  tm_facets('DAERAH') + 
  tm_shape(den_kel) +
  tm_dots(shape = 'jantina', size = 0.1) 
```


# Density of dengue cases

Aim : to calculate dengue cases per area

Steps:

- Join points `den_kel` to polygons `kel` shapefiles
- Calculate dengue points in each mukim (polygons)


```{r}
den_in_muk <- st_join(den_kel, kel, 
                      join = st_within)
glimpse(den_in_muk)
```

```{r}
count_den_mukim_mth_yr <- den_in_muk %>% 
  count(DAERAH, MUKIM, year_on, 
        month_on, JUM_JANTIN) %>% 
  ungroup()
count_den_mukim_mth_yr %>% datatable()
```


Calculate incidence of dengue per 1000 population for 

- each mukim
- year
- month

```{r}
count_den_muk_m_y_1000 <- count_den_mukim_mth_yr %>% 
  mutate(incidence_den = (n/JUM_JANTIN)*1000)
count_den_muk_m_y_1000 %>% datatable()
```

# Plot incidence of dengue for mukim

join polygon to point to enable plot polygon of Kelantan

```{r}
count_den_mukim <- den_in_muk %>% 
  count(DAERAH, MUKIM, JUM_JANTIN) %>% 
  ungroup()
count_den_mukim_1000 <- count_den_mukim %>% 
  mutate(inc_1000 = (n/JUM_JANTIN)*1000)
```


```{r}
kel_with_den <- st_join(kel, count_den_mukim_1000)
glimpse(kel_with_den)
```

Plot

```{r}
ggplot() + 
  geom_sf(data = kel_with_den, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('Dengue Incidence per 1000 population') +
  theme_bw()
```




