---
title: "Measure of Disease Occurrence and Association"
author: "Kamarul Imran Musa"
date: "11/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives

- measure of diseases occurence 
- mesure of disease association

# Prepare environment

```{r}
library(pacman)
pacman::p_load(pubh, tidyverse, epiR)
```

# pubh package

## Data

- Bernard: the effect of ibuprofen on preventing death in patients with sepsis.

```{r}
data("Bernard")
glimpse(Bernard)
```

## Transform data

Generate variables 

- variable fate with `Dead` as the reference group 
- variable treat with `Ibuprofen` as the reference group 

```{r}
Bernard %>% 
  mutate(fate = relevel(fate, ref = 'Dead'),
         treat = relevel(treat, ref = 'Ibuprofen')) %>%
  select(fate, treat) %>%
  cross_tbl(by = 'treat') %>%
  theme_pubh(2)
```

## Cross tabulation


```{r}
Bernard %>%
  contingency(fate ~ treat)
```

# epiR package

## Data

A cross sectional study investigating the relationship between dry cat food (DCF) and feline lower urinary tract disease (FLUTD) was conducted (Willeberg 1977). Counts of individuals in each group were as follows. DCF-exposed cats (cases, non-cases) 13, 2163. Non DCF-exposed cats (cases, non-cases) 5, 3349. 

```{r}
dat.v01 <- c(13, 2163, 5, 3349)
matrix(dat.v01, nrow = 2, byrow = TRUE)
```

## Calculation

```{r}
epi.2by2(dat = dat.v01, method = "cross.sectional", 
         conf.level = 0.95, units = 100, interpret = FALSE, 
         outcome = "as.columns")
```
## Manual Calculation

- Prevalence of outcome in exposed = 13/2176 = 0.00597
- Prevalence of outcome in non-exposed = 5/3354 = 0.00149
- Prevalence ratio = 0.00597/0.00149


- OE+ = 13/2163 = 0.00601
- OE- = 5/3349 = 0.00149
- OT = 18/5512 = 0.00327

- Odds Ratio = 0.00601/0.00149

- attributed prevalence in exposed = OE+ - OE- = 0.00601 - 0.00149
- attributed fraction in exposed = (OE+ - OE-)/OE+ = (0.00601 - 0.00149) / 0.00601
- attributed prevalence in population = OT - OE- = 0.00327 - 0.00149 = 0.00178
- attributed fraction in population = 0.00178 / 0.00327 = 0.54


# References

- https://cran.r-project.org/web/packages/pubh/vignettes/introduction.html 
- https://cran.r-project.org/web/packages/epiR/vignettes/epiR_measures_of_association.html
