---
title: "Outbreak Investigation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives 


# Prepare environment

```{r}
pacman::p_load(
  rio,          # file import/export
  here,         # relative filepaths 
  lubridate,    # working with dates/epiweeks
  aweek,        # alternative package for working with dates/epiweeks
  incidence2,   # epicurves of linelist data
  i2extras,     # supplement to incidence2
  stringr,      # search and manipulate character strings
  forcats,      # working with factors
  RColorBrewer, # Color palettes from colorbrewer2.org
  tidyverse,    # data management + ggplot2 graphics
  DT            #create datatable
) 
```

# Read data

Import data 

```{r}
linelist <- import(here('data',
                        'linelist_cleaned.xlsx'))
```

Check rows and columns

```{r}
glimpse(linelist)
```


# Explore data

Basic descriptive

```{r}

```

Transform data

- create a group data

```{r}
count_data <- linelist %>% 
  group_by(hospital, date_hospitalisation) %>% 
  summarize(n_cases = dplyr::n()) %>% 
  filter(date_hospitalisation > as.Date("2013-06-01")) %>% 
  ungroup()
```

View the aggregated data

```{r}
count_data %>% 
  datatable()
```

Set a reference date in case want to generate report

```{r}
data_date <- as.Date("2015-05-15")
```


```{r}
linelist <- linelist %>% 
  mutate(date_onset = as.Date(date_onset))
```



```{r}
# check range of onset dates
ggplot(data = linelist)+
  geom_histogram(aes(x = date_onset))
```

# Epidemic Curve

## using Epicurves from incidence2 package

### All

Steps:

1. Create an incidence object
2. Plot the incidence

```{r}
epi_day <- incidence(       # create incidence object
  x = linelist,             # dataset
  date_index = date_onset,  # date column
  interval = "day"          # date grouping interval
  )
```


```{r}
epi_day
```

```{r}
summary(epi_day)
```

```{r}
plot(epi_day)
```

```{r}
epi_wk <- incidence(linelist, date_onset, interval = "Monday week")
```

```{r}
plot(epi_wk)
```

### By groups

```{r}
# Create incidence object, with data grouped by age category
age_outbreak <- incidence(
  linelist,                # dataset
  date_index = date_onset, # date column
  interval = "week",       # Monday weekly aggregation of cases
  groups = age_cat,        # age_cat is set as a group
  na_as_group = TRUE)      # missing values assigned their own group

# plot the grouped incidence object
plot(
  age_outbreak,             # incidence object with age_cat as group
  fill = age_cat)+          # age_cat is used for bar fill color (must have been set as a groups column above)
labs(fill = "Age Category") # change legend title from default "age_cat" (this is a ggplot2 modification)
```


### Filtered data

```{r}
# filter the linelist
central_data <- linelist %>% 
  filter(hospital == "Central Hospital")

# create incidence object using filtered data
central_outbreak <- incidence(central_data, date_index = date_onset, interval = "week")

# plot the incidence object
plot(central_outbreak, title = "Weekly case incidence at Central Hospital")
```

### Rolling average

```{r}
rolling_avg <- incidence(                    # make incidence object
  linelist,
  date_index = date_onset,
  interval = "week",
  groups = gender) %>% 
  
  i2extras::add_rolling_average(before = 6)  # add rolling averages (in this case, by gender)

# plot
plot(rolling_avg) # faceted automatically because rolling average on groups
```

