---
title: "Practical: Outbreak Investigation"
author: "Kamarul Imran Musa"
date: "8/6/2019" 
output: 
  rmdformats::readthedown
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives 

- to load packages 
- to read the excel data
- to perform basic statistics
- to plot epidemic curve
- to plot rolling average


# Prepare environment

```{r}
pacman::p_load(
  rio,          # file import/export
  here,         # relative filepaths 
  lubridate,    # working with dates/epiweeks
  aweek,        # alternative package for working with dates/epiweeks
  incidence2,   # epicurves of linelist data
  i2extras,     # supplement to incidence2
  stringr,      # search and manipulate character strings
  forcats,      # working with factors
  RColorBrewer, # Color palettes from colorbrewer2.org
  tidyverse,
  # data management + ggplot2 graphics
  gtsummary,
  DT, #create datatable
  slider #rollingaverage
) 
```

# Read data

Import data 

```{r}
linelist <- import(here('data',
                        'linelist_cleaned.xlsx'))
```

Check rows and columns

```{r}
glimpse(linelist)
```


# Explore data

## Basic descriptive

```{r}
linelist %>% 
  select(gender, age, wt_kg, ht_cm, cough, outcome) %>%    # keep variables of interest
  tbl_summary(by = outcome) %>%  # produce summary table and specify grouping variable
  add_p()                        
```

## Transform data

- create a group data

```{r}
count_data <- linelist %>% 
  group_by(hospital, date_hospitalisation) %>% 
  summarize(n_cases = dplyr::n()) %>% 
  filter(date_hospitalisation > as.Date("2013-06-01")) %>% 
  ungroup()
```

## View aggregated data

```{r}
count_data %>% 
  datatable()
```

## Formatting date 

Set a reference date in case want to generate report

```{r}
data_date <- as.Date("2015-05-15")
```


```{r}
linelist <- linelist %>% 
  mutate(date_onset = as.Date(date_onset))
```

# Plot epidemic curve I

```{r}
# check range of onset dates
ggplot(data = linelist)+
  geom_histogram(aes(x = date_onset))
```

# Plot epidemic curve II

## Using Epicurves from incidence2 package

### All

Steps:

1. Create an incidence object
2. Plot the incidence

```{r}
epi_day <- incidence(       # create incidence object
  x = linelist,             # dataset
  date_index = date_onset,  # date column
  interval = "day"          # date grouping interval
  )
```


```{r}
epi_day
```

```{r}
summary(epi_day)
```

```{r}
plot(epi_day)
```

```{r}
epi_wk <- incidence(linelist, date_onset, 
                    interval = "Monday week")
```

```{r}
plot(epi_wk)
```

### By groups

```{r}
# Create incidence object, with data grouped by age category
age_outbreak <- incidence(
  linelist,                # dataset
  date_index = date_onset, # date column
  interval = "week",       # Monday weekly aggregation of cases
  groups = age_cat,        # age_cat is set as a group
  na_as_group = TRUE)      # missing values assigned their own group

# plot the grouped incidence object
plot(
  age_outbreak,             # incidence object with age_cat as group
  fill = age_cat)+          # age_cat is used for bar fill color (must have been set as a groups column above)
labs(fill = "Age Category") # change legend title from default "age_cat" (this is a ggplot2 modification)
```


### Epidemic curve for specific group

- Choose the group

```{r}
# filter the linelist
central_data <- linelist %>% 
  filter(hospital == "Central Hospital")
```

- Create incidence object 

```{r}
# create incidence object using filtered data
central_outbreak <- incidence(central_data, date_index = date_onset, interval = "week")
```

- Plot epidemic curve

```{r}
# plot the incidence object
plot(central_outbreak, title = "Weekly case incidence at Central Hospital")
```

### Rolling average

- using `incidence()` 
- specify a 7-day rolling average

```{r}
rolling_avg <- incidence(                    # make incidence object
  linelist,
  date_index = date_onset,
  interval = "week",
  groups = gender) %>% 
  
  i2extras::add_rolling_average(before = 6)  # add rolling averages (in this case, by gender)
```

- plot rolling average

```{r}
# plot
plot(rolling_avg) # faceted automatically because rolling average on groups
```
- if you are NOT using incidence function and calculating the rolling average directly from your data (and when there are missing dates for days with no case), for example


```{r}
daily_counts <- linelist %>% 
  count(date_hospitalisation, name = "new_cases") %>%
  ungroup()
daily_counts %>%
  datatable()
```

- then use `slide_index_dbl()` to get accurate cumulative number of cases

```{r}
rolling <- daily_counts %>% 
  mutate( #Using slide_index_dbl()
    #########################
    indexed_7day = slide_index_dbl(
        new_cases,                       # calculate on new_cases
        .i = date_hospitalisation,       # indexed with date_onset 
        .f = ~sum(.x, na.rm = TRUE),     # function is sum() with missing values removed
        .before = days(6))               # window is the DAY and 6 prior DAYS
    )
rolling %>%
  datatable()
```

- this is the plot

```{r}
ggplot(data = rolling)+
  geom_line(mapping = aes(x = date_hospitalisation, 
                          y = indexed_7day), 
            size = 1) 
```


# References

- The Epidemiologist R Handbook : https://epirhandbook.com/en/index.html 
